Long-term perspectives of social inequality in mortality among

### 2017-07-01 article.Rmd
first commit

\documentclass[a4paper,11pt,titlepage]{article}

\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}

%\usepackage{float}

\usepackage{epstopdf}
\usepackage{url}

\usepackage{natbib}

\bibliographystyle{apalike}
%\bibliographystyle{plainnat}

%\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
\newcommand{\R}{{\bf R}}

\usepackage[tablesonly]{endfloat}
%\usepackage[figuresonly]{endfloat}
%\usepackage[tablesonly]{endfloat}

\renewcommand{\efloatseparator}{\vfill}

\renewcommand{\textfraction}{0.05}
\renewcommand{\topfraction}{0.8}
\renewcommand{\bottomfraction}{0.8}
\renewcommand{\floatpagefraction}{0.75}

% patch \efloat@iwrite to use \protected@write
\makeatletter
\renewcommand\efloat@iwrite[1]{%
   \immediate\expandafter\protected@write\csname efloat@post#1\endcsname{}}
\makeatother

\author{Sören Edvinsson\footnote{email: \texttt{soren.edvinsson@umu.se}} and
Göran Broström\footnote{email: \texttt{goran.brostrom@umu.se}}
     \\
Centre for Demographic and Ageing Research ({\sc Cedar})\\
Umeå University, SE-90187 Umeå, Sweden}

\title{Long-term perspectives of social inequality in mortality among
  adults and elderly in northern Sweden 1851--2013\footnote{Paper presented
    at the
    session 'Health and Mortality in Arctic Populations'. The Ninth International
    Congress on Arctic Social Sciences,
    Umeå,
    June 8--12, 2017}}

\date{May 31, 2017}

<<setup, include = FALSE>>=
library(knitr)
opts_chunk$set(engine='R',dev='pdf',fig.path='figs/slide-',tidy=FALSE,fig.height=3.5,fig.width=8,strip.white=TRUE,cache=TRUE,echo=FALSE,message=FALSE,comment=NA,show.signif.stars = FALSE,fig.pos='ht')##$
library(knitr)
library(eha)
library(xtable)
library(skum)
@

<<readalldata>>=
## Load remote data:
con <- url("http://capa.ddb.umu.se/data/vb0.rda")
##load("data/vb0.rda")
load(con)
close(con)
vb0$hisclass[is.na(vb0$hisclass)] <- 4
vb0$hisclass <- factor(vb0$hisclass, labels = c("elite", "middle", "worker", "NA"))
##
con <- url("http://capa.ddb.umu.se/data/vb.rda")
##load("data/vb.rda")
load(con)
close(con)
##
con <- url("http://capa.ddb.umu.se/data/lisaold.rda")
##load("data/lisaold.rda")
load(con)
close(con)
##
con <- url("http://capa.ddb.umu.se/data/loTab.rda")
load(con)
close(con)
loTab$period <- factor(loTab$period)
##
vb0Tab <- aggregate(vb0[, c("event", "exposure")],
                   vb0[, c("sex", "age", "period", "cohort",
                          "urban", "civst", "hisclass")],
                   FUN = sum)
vb0Tab$age <- factor(vb0Tab$age)
@

\begin{document}

\maketitle

\begin{abstract}
A long-term perspective on the development of social
inequalities in mortality in the adult and elderly population is taken in
this paper. The area under investigation consists of the Skellefteå and Umeå regions in
the north of Sweden, and the time period is 1851--2013. The outcome is all
cause mortality and cause specific mortality for the categories
cardiovascular diseases and cancers. The analysis of inequality is based on
three social classes, reorganized from {\sc Hisclass}.

The main findings are that a consistent social gradient implying better
survival for classes with more access to economic and other resources, only
become evident in the later half of the 20th century, with lower classes
having the highest mortality. The class patterns in mortality differed
however depending on gender. Women of higher classes were comparatively
favoured in most of the studied periods, while higher class men were
disadvantaged when it comes to survival at least until the middle of the
20th century. The modern pattern of clear socio-economic differences
appeared surprisingly late. Additional and alternative analyses based on education and
income categories confirmed large inequality in mortality
between different groups during the most recent decades. There were no clear signs of qualitative differences depending on whether it was the working-active or the retired population that was analysed, but the differences according to social class was larger in the first-mentioned group. Both cardiovascular diseases and cancers conform with the pattern already observed when it comes to all-cause mortality.

\end{abstract}

\tableofcontents

\newpage

%%{\huge Här ska 'intro.tex' in !!!!!!!!!!!!!!!!!!!!!}

%%\subsection{Combining the data sets and tabulate}

\section{Introduction}

One of the central aspects of survival is social class and access to economic and other resources. Even in present-day welfare societies, social position is a strong determinant when it comes to health and mortality and the impact even seems to be increasing \citep{kuetal04, frlu07}. It has been suggested that “… social conditions have been, are and will continue to be irreducible determinants of health outcomes and therefore deserves their appellation as ‘fundamental causes’ of disease and death” \citep{liph95}. For a long time, the general view has been that socio-economic health inequalities were high in historical societies, probably higher than in modern societies. This is a reasonable assumption since these societies were in most cases characterised by very large differences between groups and classes. Knowing that access to resources provides advantages in all ways, the health advantage of higher classes ought to be obvious. \citet{ant67} and \citet{smi83} suggest that social inequality in health has passed through different historical phases. According to them, social health differences were comparatively small during the pre-transitional phase when space was a strong determinant for the spread of disease. Differences then increased during the transitional phase when mortality declined and wealthy groups used their resources to gain better health. Finally, health differences decreased again in modern low-mortality settings when instead health-related behaviour became important, resulting in marginal differences. \citet{omran82} state, in the third proposition in his theory of the epidemiological transition, that even if the class differential in mortality was maintained during the transition, the decline set in earlier and was faster among privileged groups.

Recent studies, investigating social inequality in health and mortality with micro-data, has however questioned how generalizable the assumed pattern is \citep{bevp11}. Instead, some studies suggest that the modern pattern might be a fairly recent development, although the pattern differs depending on age, gender or historical context. When it comes to the adult and elderly population results from different studies have diverged, some reporting substantial differences but many, particularly those based on Swedish data, have found surprisingly small or even reversed socio-economic differences during the first stages of the transition \citep{bedr11, edli11,edbr12}, sharply contrasting against what we know from present-day society.

Another question of the development of social inequality in mortality, concerns its relation to economic inequality. \citet{wipi09} argue that income inequality has an independent effect on mortality, separate from the direct effect of actual different access to economic resources. They argue that unequal societies perform less well when it comes to health than equal ones. Wilkinson and Pickett find a positive association between inequality and health (as well as other social conditions) at country level in the present-day economically developed world. This has initiated a vital scholarly debate and the topic has been extensively studied \citep{suka04, wado00}. How the association looked like in previous periods is unknown. This is of particular interest since the levels of economic inequality were very different from those of the recent decades. Even if it not a necessary implication, we would still assume that poorer groups were those most disadvantaged of large inequality in itself. The situation during the 19th century is uncertain, but at least during the late 19th and early 20th century both income and wealth distribution was strongly skewed in Sweden. From a couple of decades into the 20th century, income and wealth inequality continuously diminished to reach a low point at around 1980. Thereafter, economic inequality has increased substantially until the present, although far from being as high as a century ago \citep{rowa08, rowa09}. The Swedish development resembles that of other European and North American countries in its basic features \citep{rowa14}.

It thus appears as if a more consistent social health divide appeared at some point during the mortality transition, at least in Sweden. Solid empirical evidence about the process is however lacking and studies focusing on the issue are still few. There is a need for more reliable studies from different geographical and historical settings as well as according to age and gender in order to better understand the role of socio-economic conditions for health and survival over time.

% This paper is a follow-up of an earlier paper \citep{segb16b}, presented at
% the ESHD Conference in Leuven, Belgium, 2016. Here we make a quick look back
% in time for the Skellefteå region (1851--1900), and we also consider causes
% of death from the early 20th century. The {\sc Lisa} data (Swedish population registers) allow
% us to study the effect of income and education on mortality in a more
% direct way for the period 1990--2005, and to compare it with the results of the class-based analysis.

\section{Aims and questions}

In the present paper, we take a long-term perspective on the development of social inequalities in mortality in the adult and elderly population. The areas under investigation are the Skellefteå and Umeå regions in the north of Sweden, and the time period is 1851--2013 for Skellefteå and 1901--2013 for Umeå. The main question we investigate is if inequality in mortality between social classes increased among the adult and elderly population, primarily benefiting the highest social positions and how this is connected to the development of economic inequality in society. This issue is then analysed according to gender, age-group (working age vs retired) and cause of death (cardiovascular and cancer). Furthermore, we compare our results witn analyses of alternative inequality indicators, education and income for the period 
1986--2013.

% ? :
% 
% \begin{enumerate}
% \item Has inequality in mortality between social classes increased among the adult and elderly
%   population, primarily benefiting the highest social positions and how is it connected to the development of economic inequality in society?
% 
% %%\item Is the mortality transition initiated by a certain social class?
% 
% \item Are there any gender differences in the effect of social position?
% 
% \item Is social position equally important among the retired population as for those in working age?
% \item Is there a difference between classes in mortality from cardiovascular diseases and cancers respectively?
% \item Do class inequalities in survival correspond to those depending on level of education and income categories for the period 1990 on-wards?
% \end{enumerate}
% 
%

\section{The Skellefteå and Umeå regions}

% Figures~\ref{fig:mapp} and~\ref{fig:SKEUME}
% show the study areas and their position in Sweden.

% \begin{figure}
% \includegraphics[height=19cm]{myfigs/skum}
% \caption{The Skellefteå and Umeå regions.}
% \label{fig:SKEUME}
% \end{figure}

\begin{figure}
\includegraphics[height=19cm]{myfigs/sweden}
\caption{The Skellefteå and Umeå regions in Sweden.}
\label{fig:mapp}
\end{figure}

The Skellefteå and Umeå regions (Figure~\ref{fig:mapp}) are part of the county of Väster\-bot\-ten in the north of Sweden along the coast of Gulf of Bothnia. This was a remote part of the country, where communication with the rest of Sweden was difficult until the late 19th century. The economy was dominated by agriculture, making it vulnerable to harvest failures, and several severe famines affected the regions during the 1800s. One living long in memory was the famine of 1867–-1868 \citep{segb14}. During the long winters, sea communication was hindered due to the Gulf of Bothnia being frozen. The first ships during the year came to this area in May or even as late as June in some cases \citep{fah56}. Towards the end of the 19th century, the Swedish railway system reached this part of Sweden, thereby facilitating contacts with the rest of Sweden, improving the economy and mitigating the effects of harvest failures. The inland was particularly remote and isolated. During the 19th century, our regions became increasingly integrated in the same epidemiological pattern as the rest of Sweden. Diseases reached to places that earlier were spared.

The Skellefteå region until the 1950 in our dataset consists of a selection of parishes surrounding the town of Skellefteå, founded in 1845 but with a very small population during the 19th century. The data from the period after 1960 cover the Skellefteå and Malå municipalities, the same area as for the earlier period but with the addition of two more parishes. The majority of the 19th century population lived in rural villages and hamlets, getting their livelihood from agricultural production. During the 20th century, industrialisation took place, partly by the opening of mines in the in-lands. This also led to a population increase both in the town and in the rural parts, and a much more diversified economy.

Mortality was fairly low in comparison with other parts of the country and the fertility transition was late \citep{coa86}. The Skellefteå population size as defined in our data sets was 16\,473 on January 1, 1850, 41\,352 on January 1, 1900, 61\,938 on January 1, 1950 and 76\,723 at the end of the century.

The Umeå region in this study consists of Umeå urban and rural parish until 1950, and from 1960 onwards of Umeå municipality when another three parishes are included. This region has a somewhat different character from that of Skellefteå. Umeå town had a small population but was substantially larger than Skellefteå town during almost the whole studied period. It was the administrative centre in the county of Västerbotten. Schools and military regiments were placed here and the economy was more diversified compared to Skellefteå. Agriculture dominated the rural part, but there were also some foundries as well as industries for example in forestry and small-scale production. The population size as defined in our data sets was 18\,970 on January 1, 1900, 32\,900 on January 1, 1950 and 103\,970 when the century ended.

\section{Data and variables}

The recent extension in time of data at the Demographic Data Base (DDB), Umeå University (www.cedar.umu.se) by including the period between 1900 and 1950s makes it possible to investigate the long-term development of social inequality in mortality, in the present case from 1851 to 2013. The databases provide us with micro-data on a sufficiently large population and with substantial social diversity for the Skellefteå and Umeå regions \citep{wee16}. The data for the present study comes from two large population databases at the DDB. The early period until 1950 is covered by
the database Poplink, the digitisation of historical parish registers for
the Skellefteå and Umeå region \citep{wee16}. Poplink is based on linked parish records, allowing
us to reconstruct life biographies on people as long as they remained in
the region. The records are linked within but not between the regions. Data
from Poplink are accessed for the period 1851--1950.

The other large data-set is extracted from the Linnaeus database
\citep{malm10}. Linnaeus is based on different linked national population registers from
1960 to 2013 and is used within the ageing program at CEDAR, Umeå
University. The study period from 1960 to 2000 are constructed from censuses every fifth
year 1960--1990, with additional information on deaths from National Board of Health and Welfare. For the period 2002--2013 we use the information from the yearly population registers from Statistics Sweden (LISA) together with death information from National Board of Health and Welfare.

Information on individuals are anonymised and the two databases are not
linked. This stops us from following individuals present
in both databases throughout their lives, thus they are treated separately. This
also makes it impossible to add information on individuals in the Linnaeus
database from what we potentially could find in Poplink, for example family
background or occupations earlier in life.

In the data-set analysed here, all individuals 40 years and older ever being resident in either
of the regions are included. The data file contains the variables on social class ({\sc Hisclass}) sex, urban/rural residence, birth date, death date, first and last date of observation and type of
entrance/exit. Total number of persons are 84\,599 leading to 35\,001 deaths in Poplink and 197\,864 persons leading to 66\,297 deaths in the Linnaeus database. See Figure~\ref{fig:lexis2}.

<<lexis2,fig.cap="The sampling frame.", fig.height = 5>>=
plot(c(1851, 1851, 1951, 1951, 1851), c(40, 90, 90, 40, 40),
     type = "l", xlim = c(1851, 2014), lwd = 2, col = "red",
     xlab = "Year", ylab = "Age", axes = FALSE, ylim = c(0, 90))
axis(1, at = c(1851, 1901, 1951, 1961, 1981, 2014), las = 2)
axis(2, las = 2, at = c(0, 40, 65, 90))
axis(4, las = 2, at = c(0, 40, 65, 90))
box()
abline(h = 0)
lines(c(1961, 1961, 2014, 2014, 1961), c(40, 90, 90, 40, 40),
      col = "blue", lwd = 2)
#
n.rec <- NROW(vb0[vb0$dataset == "poplink", ] )
n.pl <- with(vb0[vb0$dataset == "poplink", ], length(unique(id)))
n.d <-  with(vb0[vb0$dataset == "poplink", ], sum(event))
text(1900, 70, "PopLink", cex = 2, col = "red")
text(1899, 60, paste(n.rec, "records"))
text(1900, 55, paste(n.pl, "persons"))
text(1899.5, 50, paste(n.d, "deaths"))
#
n.rec <- NROW(vb0[vb0$dataset != "poplink", ] )
n.pl <- with(vb0[vb0$dataset != "poplink", ], length(unique(id)))
n.d <-  with(vb0[vb0$dataset != "poplink", ], sum(event))

text(1985, 70, "Linnaeus", cex = 2, col = "blue")
text(1984, 60, paste(n.rec, "records"))
text(1985.2, 55, paste(n.pl, "persons"))
text(1985.2, 50, paste(n.d, "deaths"))
@

\subsection{Presence periods}

Differences in available information in the two datasets as well as within the Linnaeus data makes it necessary to apply different approaches when it comes to identification of presence periods within our
regions. The Poplink data provides us with exact dates or at least year of
start and exit of the presence of all individuals in our data, allowing us to have full and continuous control of the de jure population. This is however not the case with the Linnaeus database. For the period 1960--2000 we use presence in the censuses 1960--1990 together with the yearly population registers 1986--2000 and information on deaths from the National Board of Health and Welfare to identify the observation periods in our analyses. There is however no full information on what happened in-between the censuses, for example the exact date of exit for persons that moved between two censuses. We have solved this by considering everyone as resident in our regions during the complete five-year period between censuses if the person is missing in the later census. This overestimates their presence in our regions, but should not bias the results when it comes to socioeconomic differences. We furthermore miss those that both moved in and moved out or vice versa between censuses. Exit dates are exact if they refer to death, and in these cases we consider them as constantly present from the last census to the death data even if the death took place in another part of Sweden.

For the last period, 2002--2013, we use the yearly population registers (LISA data), together with information on deaths from National Board of Health and Welfare. This data allows good control of the presence periods in the studied regions.

\subsection{Social class}

In this paper we analyse mortality differences according to social class based on the classification scheme {\sc Hisclass} \citep{lema11}. The only type of socio-economic information available both in Linnaeus and Poplink is occupations. Thus, analysing according to social class is apart from being a scientific choice, also a practical one. Using occupational data is the only and obvious alternative for making a longitudinal study of socio-economic mortality differences with consistent classification schemes. Social position is treated as a time-dependent variable.

The availability to information on occupations varies over time and between sources. Occupations in Poplink are coded into HISCO \citep{lema02} and the original DDB coding system. The censuses 1960--1990 include occupational codes based on the system used in Sweden at that time, NYK (Nordisk Yrkesklassificering). The yearly population registers (LISA) do not report occupations until 2001, thus there is a window of 11 years from 1991 to 2001 where we lack information. In the censuses, we have solved this by extending the periods for which we consider the observed social class to be valid. Thus, the social class in the census 1990 is considered to be kept until 2001. This strategy is also used for other cases of missing social class. Furthermore, we consider the last social position before retirement to be kept until death or last observation for those above 65. For the Linnaeus data, we use a script recoding HISCO into {\sc Hisclass}, provided to us by Jonas Helgertz, Lund University.

The use of different data sets, forces us to apply different approaches towards defining the social position in some respects. A problem is that Poplink usually only provides the occupation for the head of household, thus under-estimating female working-force participation as well as that of adult children residing with their parents \citep{vik10}. In this study and according to the research questions in the present paper, we have chosen to categorise wife according to the position of the head of household, usually the husband/father, assuming that the family shares the same socio-economic position. This strategy makes the problem of under-recording less problematic.

Female labour is much better covered in the Linnaeus database. A complication is however that it is  difficult to define households in the same way as in Poplink. Extra-marital cohabitation, connected to the Second Demographic Transition \citep{van87, van03}, became common and female labour force participation developed as the norm in Sweden during this period. The occupation of the husband as signifier of social class became, if not obsolete, at least less relevant. All included persons are signified by their individual occupation for this period. Still, the results for the different periods are not completely comparable and makes the interpretation of the development for women more shaky.

{\sc Hisclass}, the classification system used here, is a "\ldots HISCO-based
historical international social class scheme" \citep{lema11, lema02}. The different classes in {\sc Hisclass} represent distinct categories, based on whether the work is manual or non–manual, skill level, supervision and sector. Implicitly it reflects large differences in what is in the forefront of other
categorisations, as for example access to economic and other resources, status, social power et cetera. We have merged the original 12 classes in {\sc Hisclass}, based on position in the labour market, ending up with the following three and an additional fourth category with unknown class:

\begin{enumerate}
\item {\tt elite}, {\sc Hisclass} 1 and 2. Higher managers, professionals and higher white collar.
\item {\tt middle}, {\sc Hisclass} 3, 4, 5, 6 and 8. Lower managers, farmers and lower white collar.
\item {\tt worker}, {\sc Hisclass} 7, 9, 10, 11 and 12. Workers including farm workers.
%%\item Lower skilled farm workers, HISCLASS 10 and 12, Lower and unskilled farm workers
\end{enumerate}

% The procedure for
% occupations in the Linnaeus database is a little more complicated. The
% occupations in the censuses (NYK) are first transferred to the system
% later in use, SSYK (the Swedish version of ISCO88). Thereafter the
% SSYK, both those transferred from the censuses and those given in the
% population registers 2001-2013, can be re-coded into HISCLASS.

Figures~\ref{fig:hiscskumm} and \ref{fig:hiscskumw} show the distribution (per cent) of exposure
time according to class for men and women respectively. The category of workers is
fairly stable over time. The highest class has increased substantially, although from a very low level in the 19th century. The increase of academics in connection to the establishment of
Umeå University explains much of this increase. What is hidden in Figure~\ref{fig:hiscskumm} is the transfer within the middle group from a majority of farmers during the 19th century to lower white collar from the middle of the 20th century. Furthermore we observe an increase during the 1960s and partly during later decades as well in those with undefined. When first observed in the earliest included censuses, occupation is often missing. For men the proportion is somewhat higher in the age group 65 and above, i.e. those that were retired, but still low enough to be used in our analyses. Until 1950, there were however very few with missing occupation. For men in working-active age, the proportion missing does not cause any problems for our analyses.

<<hiscskumm, fig.cap = "Exposure by hisclass, men (from the top: elite, middle, workers, missing) and time period.", fig.height = 4.5>>=
xx <- with(vb0Tab[vb0Tab$sex == "male", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
yy <- round(prop.table(xx, 1) * 100, 1)
barplot(t(yy)[4:1, ], col = c("red", "blue", "green", "black"), las = 2, ylab = "Per cent")
@

<<hiscskumw, fig.cap = "Exposure by hisclass, women (from the top: elite, middle, workers, missing) and time period.", fig.height = 4.5>>=
xx <- with(vb0Tab[vb0Tab$sex == "female", ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
yy <- round(prop.table(xx, 1) * 100, 1)
barplot(t(yy)[4:1, ], col = c("red", "blue", "green", "black"), las = 2, ylab = "Per cent")
@

% <<hiscskummyoung, fig.cap = "Exposure by hisclass, men 40-65 (from the top: elite, middle, workers, missing) and time period.", fig.height = 4.5>>=
% xx <- with(vb0Tab[vb0Tab$sex == "male" & vb0Tab$age %in% c("40", "45", "50", "55", "60"), ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
% yy <- round(prop.table(xx, 1) * 100, 1)
% barplot(t(yy)[4:1, ], col = c("red", "blue", "green", "black"), las = 2, ylab = "Per cent")
% @
% 
% <<hiscskumwyoung, fig.cap = "Exposure by hisclass, women 40-65 (from the top: elite, middle, workers, missing) and time period.", fig.height = 4.5>>=
% xx <- with(vb0Tab[vb0Tab$sex == "female" & vb0Tab$age %in% c("40", "45", "50", "55", "60"), ], tapply(exposure, list(period = period, hisclass = hisclass), sum))
% yy <- round(prop.table(xx, 1) * 100, 1)
% barplot(t(yy)[4:1, ], col = c("red", "blue", "green", "black"), las = 2, ylab = "Per cent")
% @

The proportion with missing values among women is however a larger problem, see Figure~\ref{fig:hiscskumw}. Corresponding to what we found for men, the proportion without social position is very low until the middle of the 20th century. This is a reflection of how womens social position is defined for that period. Missing social class is however a real problem during the 1960s and 1970s in particular. Around 80 \% lack information for the 1960s and 50 \% for the 1970s. Since social class for this period is defined according to own occupation, the consequence of many women being outside the workforce, the proportion missing becomes very high. We have chosen not to report results for women for these two decades. For the later decades, the results are fairly complete and should be representative.

% The corresponding numbers for the age group 40--65 are shown in Figures~\ref{fig:hiscskummyoung} and 
% \ref{fig:hiscskumwyoung}. Slightly lower rates of missing occupation in comparison to Figures~\ref{fig:hiscskumm} and \ref{fig:hiscskumw} are noted.

% <<nanam, results = 'asis'>>=
% library(xtable)
% ##load("data/vb0.rda")
% ##vb0$hisclass <- factor(vb0$hisclass, labels = c("elite", "lowMan", "lowWhiteC", "worker"))
% x <- with(vb0[vb0$sex == "male", ], table(period, hisclass, useNA = "ifany"))
% x <- round(prop.table(x, 1) * 100)
% xtable(x, caption = "Hisclass by time period, men. Per cent.", label = "tab:nana", digits = 0)
% @
% 
% <<nanaw, results = 'asis'>>=
% ##library(xtable)
% ##load("data/vb0.rda")
% ##vb0$hisclass <- factor(vb0$hisclass, labels = c("elite", "lowMan", "lowWhiteC", "worker"))
% x <- with(vb0[vb0$sex == "female", ], table(period, hisclass, useNA = "ifany"))
% x <- round(prop.table(x, 1) * 100)
% xtable(x, caption = "Hisclass by time period, per cent.", label = "tab:nana", digits = 0)
% @

\subsection{Marital status}

Marital status is treated as a time-dependent variable in the data file. The statuses used are unmarried, married and dissolved marriages (without distinguishing between widowed and divorces). If no explicit status is given in the sources and partner is missing, the status has been set to unmarried.

Regarding the distribution of marital status over time, see Figure~\ref{fig:civstskum}. The figure indicates an under-registration of those married and and over-registration of previously married during the 1960s.

<<civstskum, fig.cap = "Exposure by civil status (from the top: previously married, married and never married) and time period", fig.height = 4.5>>=
xx <- with(vb0Tab, tapply(exposure, list(period = period, civst = civst), sum))
yy <- round(prop.table(xx, 1) * 100, 1)
barplot(t(yy)[3:1, ], col = c("red", "blue", "black"), las = 2, ylab = "Per cent")
@

\subsection{Cause of death}
In the present study, we analyse mortality in the two most common disease
groups in the ages we are investigating---cardiovascular diseases and
cancers. National compulsory reporting of cause of death was introduced in 1911, thus our analysis of the two disease groups starts from that year. Classification systems have however changed. Sweden did not join ICD until 1952, but the historical causes of death in the Poplink database have been coded according to the ICD10 system. In the present study, we have identified the codes relevant for cardiovascular diseases and cancers in the ICD 8, 9 and 10.

The distribution of causes of death is shown in Figure~\ref{fig:causeskum}. Notice the high proportion of other causes in the first two periods. This category include cases with unknown or unspecific causes which has implications for the cause-specific analyses (see later discussion). The total dominance of our two disease groups is apparent.

<<causeskum, fig.cap = "Causes of death by time period (top: cancer, middle: cardiovascular, bottom: other).", fig.height = 4.5>>=
vb01 <- vb0[vb0$event == 1, ]
cancer <- tapply(vb01$cancer, vb01$period, sum)
heart <- tapply(vb01$heart, vb01$period, sum)
death <- tapply(vb01$period, vb01$period, length)
other <- death - cancer - heart
ttt <- rbind(cancer, heart, other)
ttt <- prop.table(ttt, 2) * 100
barplot(ttt[3:1, -(1:2)], col = 3:1, las = 2, ylab = "Per cent")
@

\subsection{Urban vs rural residence}
In 19th and early 20th century Sweden (as in other countries), mortality was much higher in urban environments. The unsanitary and overcrowded living conditions as well as the risks for importing infections made the health risks higher in towns and cities. In the present paper, we control for this aspect by distinguishing between urban and rural residence, based on in which parish people lived.

\subsection{Periods for analysis}
We have made separate analyses for the different periods. The century 1851--1950 is divided into four periods of 25 years each. The 19th century periods represent a mainly pre-industrial society dominated by agriculture and with restricted welfare provisions. During the years 1901--1950, industrialisation started and the urban environments increased in population size. The modern welfare state began to develop during this time. For the period 1960 on-wards, there are separate analyses on decades except for the two last periods. Sweden had developed into a wealthy welfare society with low income inequality in international comparison, particularly until the 1990s. A financial crisis struck Sweden in the early 1990s, which among other things increased unemployment and changed parts of the welfare system to being less generous than before. The start of the last period (2008--2013) coincides with the international financial crisis.

\subsection{Income}
Income is based on disposable individual income from the LISA database at Statistics Sweden, grouped into quartiles. It is only analysed from 1986 to 2005.

\subsection{Education}
Education is here analysed only for the period 1990 to 2013. Years of education are given in 3 categories and is based on level of education according to categories in the LISA database at Statistics Sweden. It ranges from basic education (nine years of schooling or less) at the lowest level, secondary education (high school or similar) for the middle group and to academic degree at the highest level.

\section{Models}

The proportional hazards model in a survival analysis context is used,
allowing for adjustment for civil status and rural/urban environment. Since
the central explanatory variable, {\sc Hisclass}, do not follow the
proportionality property, the analyses are stratified with respect to the
variable, and the main results are presented graphically.

Another practical problem is the size of that data set:
It contains no less than 2\,432\,871 records of 236\,011 individuals, of
which 68\,244 are observed to die.
This makes it very inconvenient to use the nowadays so popular semi-parametric
technique known as
\emph{Cox regression} \citep{cox72,gb02}. Instead we use a parametric model
that allows for reduction of the data
to \emph{sufficient statistics} \citep{fisher22}. The idea is to keep all
relevant covariates \emph{categorical}, and then create a table with all
possible combinations of levels of the covariates. For each such
combination, the total amount of exposure and the number of deaths are
calculated.  We have two sexes, three civil statuses, ten periods, two
urban/rural groups, four hisclasses at age 40,  four hisclasses at age 60,
ten age groups, and 22 cohorts, in all 422\,400 possible
combinations. However, in this scheme there are many structural zeroes
(impossible combinations) and some random zeros, so the actual number of
combinations we end up with is 24\,326. The data set is thus reduced in
size by 99 per cent.

The parametric model we use for the baseline hazard function is piece-wise
constant on five-year intervals, so we arrive at the following proportional
hazards regression model:

\begin{equation}\label{eq:prop}
h(t; \lambda, \boldsymbol{\beta}, \textbf{x}) =
\lambda(t)e^{\boldsymbol{\beta} \mathbf{x}}, \; \; 40 < t \le 90,
\end{equation}
where

\begin{equation*}
\begin{split}
\mathbf{x} &= (x_1, \ldots, x_n) \text{\ is a vector of covariates,} \\
\boldsymbol{\beta} &= (\beta_1, \ldots, \beta_n) \text{\ is the corresponding vector of regression
parameters.} \\
\end{split}
\end{equation*}
and
\begin{equation*}
\lambda(t) = \alpha_i,\; 40 + 5(i - 1) < t \le 40 + 5i, \; i = 1, \ldots, 10.
\end{equation*}

Since all components of $\mathbf{x}$ are categorical covariates, the
estimation of \eqref{eq:prop} can be performed in two ways, first by the
original data set and a survival analysis routine handling a piecewise
constant hazard, or second, by the reduced data set (table) and
\emph{Poisson regression}. The second way is faster than the first by a
factor of size at least 100, often much more, depending on hardware, and
the point is that the two ways give exactly the same results. This is
because the table is a \emph{sufficient statistic} for \eqref{eq:prop} and
the original data set.

The analyses are performed in the \R\ environment for statistical computing
and graphics \citep{cran}, especially using the package {\tt eha} 
\citep{gb12,eha}
The proportional hazards model allow us to scrutinise the differences and
the development in more detail. We divide the analysis into two age groups,
40--64 and 65--89 years of age, the first age group representing people that
were still mainly in the workforce, and the second those that mainly were 
retired. The model controls for marital
status and whether the individual resides in the urban or rural parts of the regions. There
are no serious signs of deviation from proportionality assumption, except
regarding {\sc Hisclass}. Each period are analysed separately, but we then combine results
in order to show the development of the social patterning of mortality
during the studied period.

% We analyse both regions together but perform separate analyses for men and
% women and for the age groups 40--64 and 65--89. The model in the analyses
% include social class, sex and marital status.

% We present the results for the period 2001--2013
% separately. We suggest a heterogeneity index in order to investigate a
% possible homogenisation of mortality between social classes.

The important explanatory variable, {\tt hisclass}, is included in the
proportional hazards models as a \emph{stratification} variable. The main
reason for this choice is that the effect of {\tt hisclass} on survival is
non-proportional, that is, it varies with age. With the reduced data set
and Poisson regression, this is simply achieved by introducing an
interaction term between {\tt age} and {\tt hisclass}. Then, the main
results are presented \emph{graphically}. Tables with results from the different models are presented at the end of the article, but are not commented on in detail. The comments on the results are primarily based on the figures and focuses on the central variable in our study---{\tt hisclass}. As can be seen from the tables, most variables in the model are significant at 5 per cent level.

% There is some  lack of information in the higher
% ages around 1960--1980 (see Figures~\ref{fig:hiscskumm} and \ref{fig:hiscskumw}), because that kind of 
% information is simply missing to a large extent.

\section{The effect of {\sc Hisclass}, 1851--1950, 1961--2013}

\subsection{All causes of death}

<<colorc>>=
## Colors for plotting strata:
scol <- c("blue", rep("darkgreen", 2), "red")
## Common for all groups:

labb <- levels(vb0Tab$period)
nper <- length(labb)
#source("R/getPHglm.R")
## End 'common'
@

The cumulative
hazards at the end of the time period (65 if 40--64, 90 if 65--89) are the
main target in our analysis: These numbers are used in the process of
graphically illustrating the development over time.

%creating an \emph{heterogeneity index} (a better name is needed:
%\emph{health disparity index}, \emph{Social Inequality Mortality Index, (SIMI)}?
%Something catchy.).

\subsubsection{Ages 40--64}

%The distribution over time periods.
<<getske4060>>=
cap1w <-  "Cumulative hazards for HISCLASS by time period, ages 40--64, women."
scap1w <-  "Cumulative hazards, ages 40--64, women."
cap1m <-  "Cumulative hazards for HISCLASS by time period, ages 40--64, men."
scap1m <-  "Cumulative hazards, ages 40--64, men."
##
cap0 <- "Total hazard of dying before age 65 for a 40 year old person by HISCLASS and decade, women (top) and men (bottom)."
scap0 <- "Total hazard by HISCLASS, ages 40--64."
cap12 <- "Total hazard of dying before age 65 for a 40 year old person standardised by HISCLASS and decade, women (top) and men (bottom)."
scap12 <- "Total hazard standardised by HISCLASS, ages 40--64."
cap12log <- "Total hazard of dying before age 65 for a 40 year old person by HISCLASS and decade, women (top) and men (bottom). The last decades."
scap12log <- "Total hazard by HISCLASS, ages 40--64, last years."

>>>>

---
title: "Long-term perspectives of social inequality in mortality among
  adults and elderly in northern Sweden 1851--2013"
author: "Sören Edvinsson and Göran Broström"
date: "28 juni 2017"
output:
  bookdown::html_document2:
    toc: yes
    toc_depth: 2
    toc_float: true
    number_sections: true
bibliography: ["mybib.bib"]
biblio-style: "apalike"
link-citations: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

One of the central aspects of survival is social class and access to economic 
and other resources. Even in present-day welfare societies, social position is a 
strong determinant when it comes to health and mortality and the impact even 
seems to be increasing \citep{kuetal04, frlu07}. It has been suggested that 
"... social conditions have been, are and will continue to be irreducible 
determinants of health outcomes and therefore deserves their appellation as 
'fundamental causes' of disease and death" \citep{liph95}. For a long time, 
the general view has been that socio-economic health inequalities were high in 
historical societies, probably higher than in modern societies. This is a 
reasonable assumption since these societies were in most cases characterised by 
very large differences between groups and classes. Knowing that access to 
resources provides advantages in all ways, the health advantage of higher 
classes ought to be obvious. \citet{ant67} and \citet{smi83} suggest that social 
inequality in health has passed through different historical phases. According 
to them, social health differences were comparatively small during the 
pre-transitional phase when space was a strong determinant for the spread of 
disease. Differences then increased during the transitional phase when mortality 
declined and wealthy groups used their resources to gain better health. Finally, 
health differences decreased again in modern low-mortality settings when instead 
health-related behaviour became important, resulting in marginal differences. 
\citet{omran82} state, in the third proposition in his theory of the 
epidemiological transition, that even if the class differential in mortality was 
maintained during the transition, the decline set in earlier and was faster 
among privileged groups.
